Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-10-05 joint_inf.tex
Is this a correct summary?
i had a bit of a hard time following, but maybe this is still a little unclear? :tennis:
A bit of cleaning up of response.
Pass through.

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we provide extensive empirical results for branch length bias in this setting.
\end{abstract}

@@

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.

@@

%EM Starting here, we only talk about branch lengths and don't mention fidelities, whereas just above we say "estimate a topology and branch fidelities using this"...
%EM At least remind the reader that we're actually talking about fidelities as the variables.
%das: clarification
\subsection*{Inconsistent branch parameter estimation}
When the topology is known and fixed and we estimate only the branch parameters, we show that for almost all generating parameter values, any branch parameter estimate is consistently biased.
For the branch parameter, we use branch fidelities (discussed earlier) in all statements and proofs, though these trivially extend to branch lengths via transformation.

\begin{restatable}{theorem}{blInconsist}
\label{thm:bl_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{t} \neq t^*$ everywhere except a set of measure zero.
\end{restatable}

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
%EM Totally confused here!
%EM 1. The generating probabilities do contain x* and y*.
%EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
%das: ok, took a stab at clarifying
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

%EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
%das: ha oops! you're right
This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-4\times 10^{-2}, 3\times 10^{-3}]$.

Inference on the integrated likelihood performs as expected where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
We use L-BFGS-B when optimizing \eqref{eq:marginal_likelihood}.
The errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, we have shown there are no sets of generating parameters other than ones of measure zero that yield consistent branch length estimates given the fixed, correct topology.
In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

The case of joint inference of a phylogenetic likelihood is discussed in \citet{Goldman1990-dk}.
There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the incorrect topology attains an equal likelihood value at the maximum as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased and cannot be consistent.
Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
We render the question well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.
We are able to provide much more detail on how large branch lengths must be for general joint inference to fail to be consistent.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we show empirically that branch lengths are biased in this setting even for short branches, with bias of the same order as the branch lengths themselves.
\end{abstract}

@@

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be exactly zero.
We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
Empirically, we find areas where joint inference consistently underestimates interior branch lengths, including regions of short branch length where bias is on the same order as the branch length.

@@

\subsection*{Inconsistent branch parameter estimation}
When the topology is known and fixed and we estimate only the branch parameters, we show that for almost all generating parameter values, any branch parameter estimate is consistently biased.
For the branch parameter, we use branch fidelities (discussed earlier) in all statements and proofs, though these trivially extend to branch lengths via transformation.

\begin{restatable}{theorem}{blInconsist}
\label{thm:bl_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property that $\hat{t} \neq t^*$ everywhere except a set of measure zero.
\end{restatable}

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions (in $(x^*,y^*)$) of generating fidelities.
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the profile likelihood \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
%EM I don't feel like this "range" is a very strong statement. What if we just repeat that the bias is of the same order as the branch lengths and give an example?
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-4\times 10^{-2}, 3\times 10^{-3}]$.
showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

In contrast, inference on the integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, we have shown that the only generating parameters that yield consistent branch length estimates given the correct topology lie in a set of measure zero.
In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in topological ambiguity: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when one set of sister branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; while analytically the inconsistency occurs on a four-taxon tree when one pair of sister branches are long and the other three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that the difficulties come from insufficient sequence length, which is not the case here.

The case of joint inference of a phylogenetic likelihood is discussed in \citet{Goldman1990-dk}.
There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the incorrect topology attains an equal likelihood value at the maximum as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased and cannot be consistent.
Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
Our question, in contrast, is well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.
We are able to provide much more detail on how large branch lengths must be for general joint inference to fail to be consistent.

%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.
